NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2019-10-22 Manuscript/manuscriptPGPE.tex
Update text

Zwitterionic PE is the second most abundant glycerophospholipid in eukaryotic cells
and has been related to the diseases \cite{vance15,calzada16,patel17}.
PE and PG affect membrane protein functionality \cite{hariharan18} and bind to various proteins \cite{yeagle14}.

In NMRlipids I and II project we were looking for a MD model
which would correctly reproduce headgroup and glycerol
backbone structures and cation binding for PC lipid bilayers \cite{botan15,catte16}.
Here we extend the same goal for other than PC lipids.
Currently the focus is on PE, PG and PS bilayers and their
mixtures with PC. Experimental data with different amounts of 
added salt is now collected and presented in this manuscript.

@@

Experimental order parameters for different lipid headgroups are
collected in Fig. \ref{HGorderParameters}.

Based on superficial reading, the conclusions in the literature are roughly \\ 
1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, \\
2) glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}, \\
3) headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. \\
Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
In contrast to PC lipids (see \cite{botan15} and references therein).

Several simulations containing PE, PG and PS lipids have been published \cite{??}, \todo{List should be completed}
however, glycerol backbone and headgroup order parameters are not compared to
the experiments (based on superficial reading of literature).

@@

The headgroup and glycerol backbone C--H bond order parameter magnitudes and signs of POPE and POPG
were determined by measuring the chemical-shift resolved dipolar splittings
with a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04} and
%The corresponding order parameter signs were measured with a
S-DROSS experiments~\cite{gross97} using natural abundance $^{13}$C solid state NMR spectroscopy
as described previously \cite{ferreira13,ferreira16,NMRlipidsIVps}.
%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
%The following experimental setups were used.
\todo{The rest of the details to be written. I am not sure how much we need to repeat the NMRlipidsIVps paper.}

@@

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid in the system. The average and
the standard error of the mean were then calculated over different lipids.
Python programs that use the MDAnalysis library \cite{agrawal11,gowers16}
used for all atom simulations is available in Ref. \citenum{MATCHgit}
({\tt scripts/calcOrderParameters.py}). For united atom simulations, the trajectories
with hydrogens having ideal geometry were constructed first using either buildH program~\cite{buildH}
or ({\tt scratch/opAAUA\_prod.py}) in  Ref. \citenum{MATCHgit}, and the order parameters were
then calculated from these trajectories. This approach has been tested against trajectories
with explicit hydrogens and the deviations in order parameters are small \cite{buildH,piggot18}.
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.

@@

The glycerol backbone and $\alpha$-carbon peaks in INEPT spectra of POPE were assigned based on
previously measured POPC spectra (Fig. \ref{POPEspectra}) \cite{ferreira13}. The $\beta$-carbon peak was assigned based
on $^{13}$C chemical shift table for amines available at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm}.
The order parameters for the glycerol backbone and headgroup C--H bonds were determined
from 2D-RPDLF and S-DROSS experiments (Fig. \ref{POPEspectra}), as described previously \cite{NMRlipidsIVps}.
The POPE experiments were recorded at 310~K, where the bilayer is in liquid disordered phase \cite{??}.
\todo{Details to be checked by Tiago}.

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPCPSPEPG.eps}
  \caption{\label{HGorderParameters}
    (top) Chemical structure of different lipids
    (bottom) Headgroup and glycerol backbone order parameters measured from lipids
    with different headgroups in lamellar liquid disordered phase.
    The values and signs for POPE (310~K), POPG (298~K). POPS (298~K) \cite{NMRlipidsIVps} and POPC (300~K) \cite{ferreira13,ferreira16}
    are measured using $^{13}$C NMR. The literature values for
    DOPS with 0.1M of NaCl (303~K) \cite{browning80},
    POPG with 10nM PIPES (298~K) \cite{borle85},
    DPPG with 10mM PIPES and 100mM NaCl (314~K) \cite{wohlgemuth80}, 
    DPPE (341~K) \cite{seelig76},
    E.coliPE and E.coliPG (310~K) \cite{gally81}
    are measured using $^2$H NMR. The signs from $^{13}$C NMR are used also for the literature values.
  }
\end{figure}

@@

\subsection{Headgroup and glycerol backbone of POPE and POPG in MD simulations}
As reported previously for PC \cite{botan15} and PS \cite{NMRlipidsIVps} lipids,
the headgroup and glycerol backbone order parameters show wide variation between different force fields
for both PE and PG lipids (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
and none of the force fields reproduce all values within experimental error bars.
%The Slipid simulations were able to capture the essential differences between PC and PS lipid
%headgroups \cite{NMRlipidsIVps}, but this is not the case for PE and PG lipids.
%\todo{Berger results are not here yet, but we should mention the ring like structures
%  pointed out by T. Piggot in the blog:
The poor performance of headgroup order parameters in Berger model can be probably explained by ring like structures seen in Fig. 6 in Ref. \citenum{mukhopadhyay04},
which is a typical feature for Berger based lipid force fields containing explicit hydrogen atoms in the head group \cite{zhao08,henin09,dahlberg10}.
%}

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPG.pdf}
  \caption{\label{HGorderParametersPOPG}
    The headgroup and glycerol backbone order parameters of PG lipids
    from experiments (POPG and signs from this work and from Ref.~\citenum{borle85}, %contains 10mM of PIPES,
    DPPG with 100mM NaCl from Ref.~\citenum{wohlgemuth80},% contains 10mM PIPES and,
    and E.Coli PG results from Ref.~\citenum{gally81}).
    and simulations with different force fields.
  }
\end{figure}
\begin{figure*}[!h]
  \centering
  \includegraphics[width=18.0cm]{../Figs/struct.png}
  \caption{\label{structures}
    Overlayed snapshots from CHARMM36 simulations of different lipids which give the best agreement with experiments.
  }
\end{figure*}
Also in line with previous studies for PC \cite{botan15} and PS \cite{NMRlipidsIVps} lipids,
CHARMM36 simulations are in best agreement with experiments for headgroup and glycerol backbone
order parameters, and seem to capture all the essential differences between different headroups
(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}).
%headgroup order parameters of PC PE, PG (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
%as previously observed also for PS lipids \cite{NMRlipidsIVps}.
In previous study, CHARMM36 predicted the more negative $\beta$-carbon order parameter and larger forking of
the $\alpha$-carbon is PS headgroup than in PC \cite{NMRlipidsIVps}.
In this work, the CHARMM36 simulations reproduce also the other essential differences in experimental
headgroup order parameters between different headgroups (Fig. \ref{HGorderParameters}) despite the inaccuracies
in individual segments:
The PE headgroup order parameters in CHARMM36 simulations are similar to the PC \cite{botan15}
and $\beta$-carbon order parameters is positive in PG headgroup, in contrast to the negative values observed in
other lipids. Therefore, we use CHARMM36 simulations to analyze the structural differences between
headgroups (Fig.~\ref{structures}). While rotation around N-C$_\alpha$-C$_\beta$-O$_\alpha$ dihedral was
significantly different in PS and PC headgroups in previous work \cite{NMRlipidsIVps}, essential differences
between PC, PE and PG are not observed here (Fig.~\ref{structures}) \todo{We need also the dihedral distributions to finish this discussion.}. \\
\todo{Why is the $\beta$-carbon order parameter of PG different to PC and PE then?}

@@

%Sodium binding to PC lipid bilayers is significantly overestimated by most simulations models \cite{catte16}.
%Also, sodium binding to PS lipids seems to be overestimated, although the
%presence of counterions complicate the comparison for negatively charged lipids \cite{NMRlipidsIVps}.
Similarly to POPG, the Lipid17 exhibits stronger sodium binding affinity
to POPE bilayers than Slipids, CHARMM36 and GROMOS-CKP (Fig. \ref{NAdensPE}).
In Lipid17 the sodium binding affinity to POPE is approximately similar as to POPC,
while in other models exhibit weaker binding to POPE (Fig. \ref{NAdensPC}).
The difference is small in Slipids and CHARMM36, while GROMOS-CKP predicts
subtantially stronger binding to POPC.
Experimental data for sodium binding to PE lipids is not available, while
simulations in agreement with electrometer data from NMR suggests that
sodium binding to POPC is weaker than in any of the simulations
here (Fig. \ref{NAdensPC})~\cite{catte16,melcr18}. Assuming that the binding
to POPE would be similar than to POPC, the sodium binding affinity to POPE
is potentially realistic in CHARMM36, Slipids, and GROMOS-CKP simulations here,
but substantially overestimated in Lipid17 simulation.
%On the other hand, the sodium binding affinity is stronger to POPE in Lipid17 simulations than to POPC in simulations with
%previous Amber lipid version, Lipid14.
\todo{The potential differences in the used ion models should be checked}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/NAdensPE.eps}
  \caption{\label{NAdensPE}
    Sodium (solid line) and choride ion density profiles along membrane normal
    from different simulations with PE lipids.
  }
\end{figure}

\clearpage
\subsection{Cation binding to PE and PG lipid bilayers}
The headgroup order parameters of PC lipids can used to measure ion binding
affinity to lipid bilayers, because their magnitude is linearly proportional
to the amount of bound charge in bilayer according to the molecular
electrometer concept \cite{seelig87,catte16}. The molecular electrometer concept
can be used also for bilayers containing PC lipids mixed with charged
lipids \cite{borle85,macdonald87,roux90,NMRlipidsIVps}.
%This is demonstrated in Figs \ref{OrderParametersWithCaCl},
%\ref{OrderParametersWithCaClBELOW1M} and \ref{OrderParameterCHANGESWithCaClBELOW1M},
%showing order parameters for PC headgroup $\alpha$ and $\beta$ carbons
%as a function of CaCl$_2$ concentration in the presence of different amounts of
%negatively charged PS or PG lipids.
%
%PC headgroup order parameters increase when negatively charged
%PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
%as expected based on electrometer concept \cite{seelig87}
%(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
%Further, the order parameters decrease with the addition
%of CaCl$_2$ and the decrease is more pronounced for systems with more
%negatively charged lipids (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
%At CaCl$_2$ concentrations ($\sim$ 50-300mM) where order parameters reach the values for pure PC,
%the Ca2+ binding presumably fully cancels the charge from negative lipids and
%overcharging occurs above these concenterations.
Based on the electrometer concept and other data is has been suggested that \cite{seelig90}
\begin{displayquote}
  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
    (ii) the free Ca$^{2+}$
    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
    charge, either due to protein or to lipid; \\
    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
    of bound Ca$^{2+}$ at neutral and charged lipids; \\
    (iv) the actual binding step can be described by a Langmuir
    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
    used to describe the chemical binding equilibrium.''}
\end{displayquote}
The electrometer concept has been very useful in evaluating ion binding affinity in simulations
against experiments, because the headgroup order parameter changes as a function of ion concentration
can be directly compared with experiments \cite{catte16,melcr18,NMRlipidsIVps}.

Calcium binding affinity to PC and PS lipid bilayers was not correctly described by
any of the standard MD simulation forced fields \cite{catte16,NMRlipidsIVps}, while
recently introduced force field with electronic continuum correction (ECC) performed better \cite{melcr18}.
The decrease of $\alpha$-carbon order parameter of PC lipids in PC:PG mixtures as a function
of calcium concentration is close to experiments CHARMM36 simulations (Fig. \ref{changesWITHCaClPG}),
but the decrease of $\beta$-carbon order parameter seems to be overestimated.
However, the $\beta$-carbon order parameter was not actually measured from these samples,
but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
The result is similar to the $\sim$200~ns simulations with PC lipids in previous work \cite{catte16}.
However, when simulation was continued for $\mu$s, the binding affinity substantially increased
and interpretation was that calcium overbinds to PC lipid in CHARMM36. Therefore, the
conclusion seems to be similar here, although the new NBfix parameters may complicate
the situation \todo{The status of NBfix parameters in these simulations should be checked.}.
%Anyway, the data presented in NMRlipids II project and in Fig. \ref{changesWITHCaClPG} together
%suggest that Calcium binding is similarly overestimated by CHARMM36 model in pure POPC bilayers and mixtures with
%POPG. The good agreement of $\alpha$ carbon would be explained by too weak dependence of its order
%parameter of bound charge
%\todo{The response of CHARMM36 to cationic surfactant against experiments \cite{scherer89} to be checked.
%  I have already ran the simulations, analysis to be done.}.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithCaClPG.eps}
  \caption{\label{changesWITHCaClPG}
    (left) The headgroup order parameters of PC from PC:PG mixtures as a function CaCl$_2$
    concentration from experiments \cite{borle85,macdonald87} and CHARMM36 simulations.
    Note that beta order parameter is calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}, not actually measured.
    (right) The headgroup order parameters of PG from PC:PG mixtures as a function CaCl$_2$
    concentration from experiments \cite{borle85} and CHARMM36 simulations.
  }
\end{figure*}

>>>>

PE and PG lipids are most common lipids in bacteria \cite{sohlenkamp16}.
Zwitterionic PE is the second most abundant glycerophospholipid in eukaryotic cells
and has been related to the diseases \cite{vance15,calzada16,patel17}.
Anionic PG lipids are less abundant, but is also proposed to be fundamental for terrestrial life \cite{furse17}.
PE and PG affect membrane protein functionality \cite{hariharan18} and bind to various proteins \cite{yeagle14}.
PE headgroup is also prone for negative membrane curvature and causes membrane fusion \cite{??}.
Therefore, the PE and PG headgroup structures play probably essential roles in 
many biological processes.

Structural details of lipid headgroups are mainly studied using NMR experiments, which
suggest that the glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, 
glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90},
and the headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. 
%Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
%In contrast to PC lipids (see \cite{botan15} and references therein).
Some attempts to resolve conformational ensembles from NMR for PC lipids have been made,
but lesser extend for other lipids \cite{??}.
Classical molecular dynamics simulations could potentially give such ensembles and therefore enable
the detailed studies of lipid headgroup behaviour in complex biomolecular systems, but current
force fields are not accurate enough to reproduce the correct conformational ensembles for PC and PS headgroups \cite{botan15,antila19}.
Several MD simulations of PE and PG lipids have been published especially in the context of modeling
inner membrane of Gram-negative bacteria \cite{??}, but glycerol backbone and headgroup structures have not
been evaluated against experiments.

Besides the structure, also ion binding may regulate biophysical activity of especially negatively
charged lipid headgroups \cite{??}. Monovalent cation (except Lithium) binding to zwitterionic PC headgroups is
very weak, while multivalent ion binding is stronger but still weak \cite{??}. The ion binding affinity
data for PE is more scarce \cite{??}, but large differences to PC would be surprising.
Negatively charged lipids are suggested to bear same cation binding constants than zwitterionic lipids, but
the amount of bound ions to negatively charged membranes would still be larger because
the concentration of cations in the vicinity of membranes would be higher  \cite{seelig90}.
%is stronger
%Based on the electrometer concept and other data is has been suggested that \cite{seelig90}
%\begin{displayquote}
%  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
%    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
%    (ii) the free Ca$^{2+}$
%    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
%    charge, either due to protein or to lipid; \\
%    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
%    of bound Ca$^{2+}$ at neutral and charged lipids; \\
%    (iv) the actual binding step can be described by a Langmuir
%    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
%    used to describe the chemical binding equilibrium.''}
%\end{displayquote}
On the other hand, anionic PS lipids are proposed chelate with calcium ions \cite{??}.
In simulations, the cation binding affinity to PC and PS membranes is typically overestimated \cite{catte16,antila19},
which can be improved by applying the ECC to the partial charges of the force fields \cite{melcr18,melcr19}.

Here, we use open collaboration and order parameters of glycerol backbone and headgroup
to evaluate the accuracy of PE and PG heagroup structures, and the cation binding affinity
to anionic membranes containing PG lipids in the current MD simulation force fields.
The force field giving the best description for glycerol backbone and headgroup structures
of PC, PS, PG and PE headgroups (CHARMM36) reproduces the essential differences in order parameters
between these headgroups, and therefore enables the analysis of structural differences between the
headgroups.

@@

@@

The headgroup and glycerol backbone C--H bond order parameter magnitudes and signs of POPE and POPG
were determined by measuring the chemical-shift resolved dipolar splittings
with a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04} and
%The corresponding order parameter signs were measured with a
S-DROSS experiments~\cite{gross97} using natural abundance $^{13}$C solid state NMR spectroscopy
as described previously \cite{ferreira13,ferreira16,antila19}.
%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
%The following experimental setups were used.
\todo{The rest of the details to be written. I am not sure how much we need to repeat the NMRlipidsIVps paper.}

@@

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid in the system. The average and
the standard error of the mean were then calculated over different lipids.
Python programs that use the MDAnalysis library \cite{agrawal11,gowers16}
used for all atom simulations is available in Ref. \citenum{MATCHgit}
({\tt scripts/calcOrderParameters.py}). For united atom simulations, the trajectories
with hydrogens having ideal geometry were constructed first using either buildH program~\cite{buildH}
or ({\tt scratch/opAAUA\_prod.py}) in  Ref. \citenum{MATCHgit}, and the order parameters were
then calculated from these trajectories. This approach has been tested against trajectories
with explicit hydrogens and the deviations in order parameters are small \cite{buildH,piggot17}.
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.

@@

The glycerol backbone and $\alpha$-carbon peaks in INEPT spectra of POPE were assigned based on
previously measured POPC spectra (Fig. \ref{POPEspectra}) \cite{ferreira13}. The $\beta$-carbon peak was assigned based
on $^{13}$C chemical shift table for amines available at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm}.
The order parameters for the glycerol backbone and headgroup C--H bonds were determined
from 2D-RPDLF and S-DROSS experiments (Fig. \ref{POPEspectra}), as described previously \cite{antila19}.
The POPE experiments were recorded at 310~K, where the bilayer is in liquid disordered phase \cite{??}.
\todo{Details to be checked by Tiago}.

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPCPSPEPG.eps}
  \caption{\label{HGorderParameters}
    (top) Chemical structure of different lipids
    (bottom) Headgroup and glycerol backbone order parameters measured from lipids
    with different headgroups in lamellar liquid disordered phase.
    The values and signs for POPE (310~K), POPG (298~K). POPS (298~K) \cite{antila19} and POPC (300~K) \cite{ferreira13,ferreira16}
    are measured using $^{13}$C NMR. The literature values for
    DOPS with 0.1M of NaCl (303~K) \cite{browning80},
    POPG with 10nM PIPES (298~K) \cite{borle85},
    DPPG with 10mM PIPES and 100mM NaCl (314~K) \cite{wohlgemuth80}, 
    DPPE (341~K) \cite{seelig76},
    E.coliPE and E.coliPG (310~K) \cite{gally81}
    are measured using $^2$H NMR. The signs from $^{13}$C NMR are used also for the literature values.
  }
\end{figure}

@@

\subsection{Headgroup and glycerol backbone of POPE and POPG in MD simulations}
As reported previously for PC \cite{botan15} and PS \cite{antila19} lipids,
the headgroup and glycerol backbone order parameters show wide variation between different force fields
for both PE and PG lipids (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
and none of the force fields reproduce all values within experimental error bars.
%The Slipid simulations were able to capture the essential differences between PC and PS lipid
%headgroups \cite{antila19}, but this is not the case for PE and PG lipids.
%\todo{Berger results are not here yet, but we should mention the ring like structures
%  pointed out by T. Piggot in the blog:
The poor performance of headgroup order parameters in Berger model can be probably explained by ring like structures seen in Fig. 6 in Ref. \citenum{mukhopadhyay04},
which is a typical feature for Berger based lipid force fields containing explicit hydrogen atoms in the head group \cite{zhao08,henin09,dahlberg10}.
%}

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPG.pdf}
  \caption{\label{HGorderParametersPOPG}
    The headgroup and glycerol backbone order parameters of PG lipids
    from experiments (POPG and signs from this work and from Ref.~\citenum{borle85}, %contains 10mM of PIPES,
    DPPG with 100mM NaCl from Ref.~\citenum{wohlgemuth80},% contains 10mM PIPES and,
    and E.Coli PG results from Ref.~\citenum{gally81}).
    and simulations with different force fields.
  }
\end{figure}
\begin{figure*}[!h]
  \centering
  \includegraphics[width=18.0cm]{../Figs/struct.png}
  \caption{\label{structures}
    Overlayed snapshots from CHARMM36 simulations of different lipids which give the best agreement with experiments.
  }
\end{figure*}
Also in line with previous studies for PC \cite{botan15} and PS \cite{antila19} lipids,
CHARMM36 simulations are in best agreement with experiments for headgroup and glycerol backbone
order parameters, and seem to capture all the essential differences between different headroups
(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}).
%headgroup order parameters of PC PE, PG (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
%as previously observed also for PS lipids \cite{antila19}.
In previous study, CHARMM36 predicted the more negative $\beta$-carbon order parameter and larger forking of
the $\alpha$-carbon is PS headgroup than in PC \cite{antila19}.
In this work, the CHARMM36 simulations reproduce also the other essential differences in experimental
headgroup order parameters between different headgroups (Fig. \ref{HGorderParameters}) despite the inaccuracies
in individual segments:
The PE headgroup order parameters in CHARMM36 simulations are similar to the PC \cite{botan15}
and $\beta$-carbon order parameters is positive in PG headgroup, in contrast to the negative values observed in
other lipids. Therefore, we use CHARMM36 simulations to analyze the structural differences between
headgroups (Fig.~\ref{structures}). While rotation around N-C$_\alpha$-C$_\beta$-O$_\alpha$ dihedral was
significantly different in PS and PC headgroups in previous work \cite{antila19}, essential differences
between PC, PE and PG are not observed here (Fig.~\ref{structures}) \todo{We need also the dihedral distributions to finish this discussion.}. \\
\todo{Why is the $\beta$-carbon order parameter of PG different to PC and PE then?}

@@

%Sodium binding to PC lipid bilayers is significantly overestimated by most simulations models \cite{catte16}.
%Also, sodium binding to PS lipids seems to be overestimated, although the
%presence of counterions complicate the comparison for negatively charged lipids \cite{antila19}.
Similarly to POPG, the Lipid17 exhibits stronger sodium binding affinity
to POPE bilayers than Slipids, CHARMM36 and GROMOS-CKP (Fig. \ref{NAdensPE}).
In Lipid17 the sodium binding affinity to POPE is approximately similar as to POPC,
while in other models exhibit weaker binding to POPE (Fig. \ref{NAdensPC}).
The difference is small in Slipids and CHARMM36, while GROMOS-CKP predicts
subtantially stronger binding to POPC.
Experimental data for sodium binding to PE lipids is not available, while
simulations in agreement with electrometer data from NMR suggests that
sodium binding to POPC is weaker than in any of the simulations
here (Fig. \ref{NAdensPC})~\cite{catte16,melcr18}. Assuming that the binding
to POPE would be similar than to POPC, the sodium binding affinity to POPE
is potentially realistic in CHARMM36, Slipids, and GROMOS-CKP simulations here,
but substantially overestimated in Lipid17 simulation.
%On the other hand, the sodium binding affinity is stronger to POPE in Lipid17 simulations than to POPC in simulations with
%previous Amber lipid version, Lipid14.
\todo{The potential differences in the used ion models should be checked}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/NAdensPE.eps}
  \caption{\label{NAdensPE}
    Sodium (solid line) and choride ion density profiles along membrane normal
    from different simulations with PE lipids.
  }
\end{figure}

\clearpage
\subsection{Cation binding to PE and PG lipid bilayers}
The headgroup order parameters of PC lipids can used to measure ion binding
affinity to lipid bilayers, because their magnitude is linearly proportional
to the amount of bound charge in bilayer according to the molecular
electrometer concept \cite{seelig87,catte16}. The molecular electrometer concept
can be used also for bilayers containing PC lipids mixed with charged
lipids \cite{borle85,macdonald87,roux90,antila19}.
%This is demonstrated in Figs \ref{OrderParametersWithCaCl},
%\ref{OrderParametersWithCaClBELOW1M} and \ref{OrderParameterCHANGESWithCaClBELOW1M},
%showing order parameters for PC headgroup $\alpha$ and $\beta$ carbons
%as a function of CaCl$_2$ concentration in the presence of different amounts of
%negatively charged PS or PG lipids.
%
%PC headgroup order parameters increase when negatively charged
%PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
%as expected based on electrometer concept \cite{seelig87}
%(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
%Further, the order parameters decrease with the addition
%of CaCl$_2$ and the decrease is more pronounced for systems with more
%negatively charged lipids (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
%At CaCl$_2$ concentrations ($\sim$ 50-300mM) where order parameters reach the values for pure PC,
%the Ca2+ binding presumably fully cancels the charge from negative lipids and
%overcharging occurs above these concenterations.
The electrometer concept has been very useful in evaluating ion binding affinity in simulations
against experiments, because the headgroup order parameter changes as a function of ion concentration
can be directly compared with experiments \cite{catte16,melcr18,antila19}.

Calcium binding affinity to PC and PS lipid bilayers was not correctly described by
any of the standard MD simulation forced fields \cite{catte16,antila19}, while
recently introduced force field with electronic continuum correction (ECC) performed better \cite{melcr18}.
The decrease of $\alpha$-carbon order parameter of PC lipids in PC:PG mixtures as a function
of calcium concentration is close to experiments CHARMM36 simulations (Fig. \ref{changesWITHCaClPG}),
but the decrease of $\beta$-carbon order parameter seems to be overestimated.
However, the $\beta$-carbon order parameter was not actually measured from these samples,
but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
The result is similar to the $\sim$200~ns simulations with PC lipids in previous work \cite{catte16}.
However, when simulation was continued for $\mu$s, the binding affinity substantially increased
and interpretation was that calcium overbinds to PC lipid in CHARMM36. Therefore, the
conclusion seems to be similar here, although the new NBfix parameters may complicate
the situation \todo{The status of NBfix parameters in these simulations should be checked.}.
%Anyway, the data presented in NMRlipids II project and in Fig. \ref{changesWITHCaClPG} together
%suggest that Calcium binding is similarly overestimated by CHARMM36 model in pure POPC bilayers and mixtures with
%POPG. The good agreement of $\alpha$ carbon would be explained by too weak dependence of its order
%parameter of bound charge
%\todo{The response of CHARMM36 to cationic surfactant against experiments \cite{scherer89} to be checked.
%  I have already ran the simulations, analysis to be done.}.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithCaClPG.eps}
  \caption{\label{changesWITHCaClPG}
    (left) The headgroup order parameters of PC from PC:PG mixtures as a function CaCl$_2$
    concentration from experiments \cite{borle85,macdonald87} and CHARMM36 simulations.
    Note that beta order parameter is calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}, not actually measured.
    (right) The headgroup order parameters of PG from PC:PG mixtures as a function CaCl$_2$
    concentration from experiments \cite{borle85} and CHARMM36 simulations.
  }
\end{figure*}
